Probability of causal contacts between interstellar civilizations through Monte Carlo simulations

### 2020-05-26 submit_v03/manuscript.tex
texto

\newcommand{\ceti}{CCN}
\newcommand{\cetis}{CCNs}
\newcommand{\hs}{\textsc{hearsay}}

@@

%%% S E C T I O N - - - - - - - - - - - - - - - - - - - - - - -
\section{Motivations}\label{S_motivations}

@@

Indeed, for the number of years projects for the Search of
Extraterrestrial Intelligences (SETI) have been working, no signal has
been received using the technical possibilities within their search
criteria \citep{tarter_search_2001}.

@@

less prone to a formal study to define searching strategies or to

@@

The optimistic estimations from the Drake equation also contrast with
the results from the SETI initiatives.

@@

Notably, despite decades of effort not a single evidence for
intelligent signals has been detected.

@@

time since the Galaxy formed and examine trajectories for the

@@

and their contacts have also been explored by several authors

@@

be modified with time, it also depend on the definition of a number of
parameters which are unknown or uncertain.

@@

\citet{grimaldi_signal_2017} consider a statistical model for the

@@

In a recent work, \citep{balbi_impact_2018} use a statistical model to
analyse the occurrence of causal contacts between civilisations in the
Galaxy.
%
The author highlights the effect of evolutionary processes when
attempting to estimate the number of communicating civilisations that
might be in causal contact with an observer on the Earth.
%
\citet{cirkovic_temporal_2004} also emphasise the lack of temporal
structure in the Drake equation.
%
\citet{bloetscher_using_2019} uses a probabilistic approach, though
still motivated by the Drake equation, to obtain a probabilistic
measure for the number of civilisations in the Galaxy.
%
To that end, the authors do Monte-Carlo Markov Chains over each factor
of the Drake equation, and combine the means to reach a probabilistic
result.
%
It is worth mentioning that the author proposes a log-normal target
distribution to compute the posterior probabilities.
%
This study concludes that there is a tiny probability for Earth to
contact another extraterrestrial civilization.    
%
\citet{smith_broadcasting_2009} uses an analytical model to gauge the
probabilities of contact between two randomly located civilisations,
and the waiting time for the first contact, assuming a fixed maximum
broadcasting distance.
% 
Adopting a novel approach, \citet{balbi_impact_2018} investigate the
chance of communicating civilisations making causal contact within a
volume surrounding the location of the Earth, using a statistical
model to compute the occurrence of causal contacts.
%
The author stresses the fact that the causal contact requirement
involves the distance between civilisations, their lifespan and their
times of appearance. This is important since the time it takes the
light to travel across the Galaxy is much lesser than its lifetime.
%
\citet{balbi_impact_2018} fix the total number of civilisations and
explore the parametric space that comprises three parameters, namely,
the distance to the Earth, the time of appearance and the lifespan of
the communicating civilization.
%
Each of these three variables are drawn from a random distribution.
%
The distribution of the distances results from a uniform model for the
positions of civilizations within the plane of the Galaxy.
%
For the distribution of the characteristic time of appearance, the
author explores exponential and truncated Gaussian distributions.
%
Moreover, for the distribution of the lifespans, the author chooses an
exponential distribution.
%
\citet{balbi_impact_2018} finds that the fraction of communicating
civilisations vary with the choice of the statistical distribution for
the time of appearance. 
%
For all analysed distributions, the fraction corresponding to a mean
lifetime of 10$^9$ years and for a maximum radius for the detection of
the signal of 1000~ly, is roughly 0.1.
%
In another related approach, \citet{vukotic_astrobiological_2012}
propose a probabilistic cellular automata (PCA) modelling. 
%
In this framework, a complex system is modelled by a lattice of cells
which evolve at discrete time steps, according to transition rules
that take into account for each cell the states of its neighbour
cells.
%
The authors implement a PCA model to a network of cells which
represent life complexity on a 2-dimensional region resembling the
Galactic Habitable Zone (GHZ),  an annular ring set between a minimal
radius of 6 kpc and a peak radius of 10 kpc. 
%
Within this framework, \citet{vukotic_astrobiological_2012} also make
several Monte Carlo simulations and analyse ensemble-averaged results.
%
The work aims at analysing the evolution of life, although it does not
account for the network of causal contacts among
technological civilisations.

@@

In this work we address the problem of the temporal and spatial
structure of the distribution of communicating civilisations.
%
We make no asumptions about, e.g., the origin of life, the development
of intelligence or the formation of habitable planets according to
stellar type.
%
Instead, we assume monoparametric function families to model the
appearance of points in the disk of the Galaxy and over the time,
which we call \textit{nodes}.
%
These nodes represent the locations of ideal intelligent agents that
are able to receive and emit signals with perfect efficiency in all
directions.
%
Then we analyse the network of causally connected nodes limited to
have a maximum separation representing the maximum distance a signal
can travel with an intensity above a fixed threshold.
%
A node is causally connected to other node if it is within their light
cone.
%
We adopt as a definition that a light cone is the region in spacetime
within the surface generated by the light emanating from a given point
in space in a given period of time.
%
For example, two nodes separated by 100 light years which appear with
a difference in time of 80 years and last 50 years each, will be
causally connected for 30 years.
%
This connection is not bidirectional.
%
In fact, the second node will see the first one for 30 years, but the
first node will never acknowledge the existence of the second.
%
This simple tought experiment exemplify the importance of the time
variable to analyse the structure of connections.
%
The model can be described with three parameters, namely, the mean separation
between the appearance of nodes, the mean lifetime of the nodes, and
the maximum separation to allow an effective causal contact.
%
The model presented here comprises 
the stochastic networks of constrained causally connected nodes,
together with the three parameters and additional hypotheses.
%
Hereafter, we refeer to this model as the C3Net model, standing for
''Constrained Causally Conencted Network``.

We abandon the frequentist approach of the Drake equation to
compute the number of civilisations, providing its statistical
distribution instead.
%
This is an exploratory analysis that aims at developing a numerical
tool to discuss the different scenarios based on statistical
heuristics.
%
With this, we avoid the several theoretical problems
summarised by the Drake equation factors.
%
The approach proposed here should be considered as a compromise
between the uncertainties of the frequentist estimations and the detailed
recipes required on the numerical simulations.
%
We provide a framework to explore, through a suite of
numerical simulations, a parameter space of three unknown observables.
%
This allows to discuss possible scenarios and their consequences in
terms of the probability of making contacts.
%
The method we use for simulating a stochastic process is an
approximation that allows to study the behaviour of complex systems,
by considering a sequence of well defined discrete events.
%
The simulation is carried out by following all the variables that
describe and constitute the state of the system, and the evolution of
the process is described as a set of changes in that state.
%
In this context, an event produces a specific change in the state that
can be triggered by random variables that encode the stochastic nature
of the physical phenomenon.
%
For example, when a new contact is produced between two entities in
the simulated galaxy, the numbers of active communication lines and of
communicated nodes does change.
%
The process involves following the changes on the state of the system,
defining the initial and final states.
%
This is done by defining a method that allows to keep track of the
time progress in steps, and maintaining a list of relevant events.
%
In Sec.~\ref{S_methods} we introduce the methods and discuss the
candidate distributions for the statistical aspects of the times
involved in the communication process.
%
We present our results in Sec~\ref{S_results}, with special emphasis
on the statistical distributions of the duration of causal contacts in
one or both directions and the distribution of time intervals for the
waiting of the first contact.
%
This quantities are considered as a function of the three simulation
parameters.
%
In Sec.~\ref{S_discussion} we discuss our results and future research
directions.   
%
In this work we use the International System of units recommended by
the International Astronomial Union, where the simbol for years is
''a''.
%}}}

\begin{figure}[!t]
   \centering
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   \caption{
   %
Space--time diagrams showing a schematic representation of the
different stages in the development of causal contact node (\ceti{}).
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (left vertex), following an "A" type event
("Awakening") in which the node acquires the communication capability.
%
The sphere of first contact (SFC) is a sphere centered on the emitter
that grows until its radius reaches the D$_{\rm max}$ distance, in which
the power of a signal would equal the detectability threshold.
%
In the Figure this sphere is represented by the left triangle of the
region $R$.
%
The surface of last contact (SLC) is another sphere that grows from a
"D" event ("Doomsday").
%
The region which is causally connected to the emitter is then limited
by these two spheres, and has the shape of a sphere or of a spherical
shell, depending on the time.
%
The temporal intervals for
the communication between two \cetis{} are represented in he
panel (b).
%
The receiver \ceti{} E$_1$ can listen signals from the emiter \ceti{} E$_2$,
from the "Contact" event (t=C$_{12}$) up to the "Blackout" event (t=B$_{12}$).
%
Similarly, 
the receiver \ceti{} E$_2$ can listen signals from the emiter \ceti{} E$_1$,
from t=C$_{21}$ up to t=B$_{21}$. 
%
   } \label{F_scheme}
\end{figure}

@@

Simulations are suitable tools to analyse systems that evolve with time and involve randomness.
%
An advantage of a numerical approach is that usually require fewer assumptions and simplifications, and can be applied to systems where analytical models are not easily found.
%
In particular, a suitable tool to model complex stochastic processes through the changes in the state of a system is the discrete--event (hereafter, DE) simulation approach.
%
A system described with the DE paradigm is characterised by a set of actors and events.
%
Actors interact causally through a series of events on a timeline and process them in chronological order \citep{ptolemaeus_system_2014, chung_simulation_2003, ross_simulation_2012}.
%
Each event modify the variables that define the process, producing the corresponding change in the state of the simulated system.
%
This method is well suited for the particular case of the diffusion of intelligent signals in the galaxy and allows to explore several models easily.
%
We simulate the statistical properties of a set of points in space and time that have a causal connection at light speed, dubbed ‘’Causal Contact Nodes’’ (hereafter, \ceti{}).
%
We chose this generic name in order to stress the fact that in this analysis 
only the causal contact is considered, independently of any broadcasting or lookout activity.
%
The system we propose is ideal, in the sense that it considers the special case of a fully efficient node that emits and receives isotropically.
%
A causal contact node, considered as a broadcasting station that has the ability to detect signals through an active lookout program, could have a lesser than one efficiency factor, which is not included in the current analysis.
%
Also, it is worth mentioning that this is a general approach, and not necessarily a \ceti{} is the host of an intelligent civilization.
%
For instance, it can be associated with a planet where life has developed, became intelligent, reached the skills required to find the right communication channel, sustain a search and establish a contact.
%
Alternative message processing entities could be considered, for example interstellar beacons where life has ceased to exist but continue with its emission, or communication stations established by probes or left by intelligent beings \citep[see, e.g., ][]{peters_outer_2018, barlow_galactic_2013}.
%
In principle, these strategies could affect our results since it would be easier to configure a cluster of \cetis{} that spread in time.
%
However, we do not consider these speculative alternatives at this point.
%
For the purpose of the present analysis, only the communication capability is relevant, since we study the causal contacts between the locations.
%
The system is defined by a number of actors that represent \cetis{} and appear at different instants in time, generating events that produce meaningful changes in the variables that describe the system, i.e., in the arrangement of \cetis{} and their network of causal contacts.
%
For example, the appearance of a new \ceti{} in a region filled with a signal emitted by another node, will increase the number of active \cetis{} and the number of pairs of \cetis{} in causal contact.
%
Assuming some simple hypotheses, the discrete events method can be performed taking into account a small number of variables, which allow to analyze the variation of the results in the model parameter space.

In what follows, we outline the experimental setup adopted to estimate the probabilities of causal contacts and several derived quantities.
%
This is done in terms of three independent parameters, namely, the
mean time span between the appearance of consecutive \cetis{},
$\left<\tau_a\right>$, their mean lifetime $\left<\tau_s\right>$, and
the maximum distance a signal can be detected by another \ceti{}
($D_{\rm max}$).
%
Intuitively, the probability of the existence of causal contacts
between pairs of \cetis{} would be larger for smaller $\tau_a$
parameters, higher $\tau_s$ or higher $D_{\rm max}$ parameter values.
%
We also propose theoretical distribution functions for both the lifespans ($\tau_s$) and the number of \cetis{} per unit time \citep{maccone_evolution_2014, Sotos_biotechnology_2019}.
%
The later is related to the time span between the appearance of consecutive \cetis{} (since when $\left<\tau_a\right>$ is shorter, it produces a greater density of \cetis{}).
%
The analytical expresiones for these distributions are set to a fixed law, as discussed in Sec.~\ref{SS_PDF_shape}.

@@

The temporal structure that emerges from the experimental setup implies that, as a fundamental property of the simulations, a causal connection can be produced without 
requiring that the two nodes are active at the same time.
%
This property of the system arises as a consequence of the large spatial and temporal scales, where a message is transmitted at the (relatively small) light speed.
%
Although a \ceti{} could be active for a large enough period to transmit a message at large distances, the limited power of the message and the dilution that depends on the squared distance from the source imposes a detectability limit.
%
As a consequence of this limitation and of their finite lifetime, considered as the period between the acquisition and loss of communicating capacity, each \ceti{} will fill a spherical shell region of the galaxy, limited by two concentric spherical surfaces.
%
The leading front, or surface of first contact (SFC) grows from the
central \ceti{} until it reaches the maximum distance $D_{\rm max}$.
%
Following the end of the civilization, there is still a region which is filled with the emitted signals.
%
This approach has been also considered in other statistical models \citep[e.g., ][]{smith_broadcasting_2009, grimaldi_signal_2017, Grimaldi2018}.
%
The trailing front, or surface of last contact (SLC) also grows from the central \ceti{}, with a delay with respect to the SFC equivalent to the lifetime of the node, and produces a spherical shell region.
%
Any other \ceti{} within this region will be aware of the originating node, even if it has disappeared before the time of contact.
%
This region will grow if the surface of first contact has not yet
reached the maximum distance $D_{\rm max}$, and will shrink after a
D-type event until the surface of last contact reaches $D_{\rm max}$, producing as a result the loss of all signal from the central \ceti{}.
%
In our approach, we consider a model galaxy where the width of the disk is negligible with respect to the radius of the disk.
%
In the 2D simulation only the intersection of the communicating spherical shells with the plane of the galaxy is relevant, and produce the corresponding circles or rings for the filled spheres or annular regions of the spherical shells, respectively.
%
The initially growing communicating sphere is shown over space--time diagrams, where time is represented on the horizontal direction, and space is represented in the vertical direction.
%
In the Fig.~\ref{F_scheme} the two emitters in panel (b), E$_1$ and E$_2$, reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emiter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the distance and time lag between the awakening events, the maximum distance that a signal can reach and the time period in which each emitter is active.

>>>>

\newcommand{\ceti}{C3N}
\newcommand{\cetis}{C3Ns}
\newcommand{\hs}{\textsc{hearsay}}

@@

%%% S E C T I O N - - - - - - - - - - - - - - - - - - - - - - -
\section{Introduction}\label{S_motivations}

@@

Indeed, as \citep{tarter_search_2001} pointed out, according to our current technical
capabilities for SETI, we have not received any signal yet.
%Indeed, for the number of years projects for the Search of
%Extraterrestrial Intelligences (SETI) have been working, no signal has
%been received using the technical possibilities within their search
%criteria \citep{tarter_search_2001}.

@@

less suitable to a formal study to define searching strategies or to

@@

%The optimistic estimations from the Drake equation also contrast with
%the results from the SETI initiatives.

@@

%Notably, despite decades of effort not a single evidence for
%intelligent signals has been detected.
A number of studies propose alternative formalisms for the estimation of the
likelihood of detecting intelligent signals from space.

@@

time since the Galaxy formed, and examine trajectories for the

@@

and their contacts have been explored by several authors

@@

be modified with time (like, for instance, the fraction of stars with planets), 
it also depend on the definition of a number of
parameters which are unknown or uncertain, required to carry out
a the simulations.

@@

\citet{grimaldi_signal_2017} considers a statistical model for the

@@

In a recent work, \citet{balbi_impact_2018} uses a statistical model to
analyse the occurrence of causal contacts between civilisations in the
Galaxy.
%
The author highlights the effect of evolutionary processes when
attempting to estimate the number of communicating civilisations that
might be in causal contact with an observer on the Earth.
%
\citet{cirkovic_temporal_2004} also emphasises the lack of temporal
structure in the Drake equation and, in particular, the limitations
of this expresion to estimate the required timescale of a SETI
program to succeed in the detection of intelligent signals.
%
Despite its shotcomming for a solid mathematical foundation of SETI, 
the Drake equation has also motivated efforts to produce formal
statistical results.
%
Recently, \citet{bloetscher_using_2019} considers a Bayesian approach,
still motivated by the Drake equation, to estimate
the number of civilisations in the Galaxy.
%
To that end, the authors employ Monte Carlo Markov Chains over each factor
of the Drake equation, and combine the means to reach a probabilistic
result.
%
It is worth mentioning that the author proposes a log-normal target
distribution to compute the posterior probabilities.
%
This study concludes that there is a tiny probability for Earth to
contact another extraterrestrial civilization.    
%
\citet{smith_broadcasting_2009} uses an analytical model to gauge the
probabilities of contact between two randomly located civilisations,
and the waiting time for the first contact, assuming a fixed maximum
broadcasting distance.
% 
Different statistical modelings have been also proposed.
%
A notable example of this es the work by \citet{balbi_impact_2018}, who investigates the
chance of communicating civilisations making causal contact within a
volume surrounding the location of the Earth, using a statistical
model to compute the occurrence of causal contacts.
%
The author stresses the fact that the causal contact requirement
involves the distance between civilisations, their lifespan and their
times of appearance.
%
This is important since the time it takes the
light to travel across the Galaxy might be much lesser than its lifetime.
%
\citet{balbi_impact_2018} fixes the total number of civilisations and
explore the parametric space that comprises three parameters, namely,
the distance to the Earth, the time of appearance and the lifespan of
the communicating civilization.
%
Each of these three variables are drawn from a random distribution.
%
The distribution of the distances results from a uniform model for the
positions of civilizations within the plane of the Galaxy.
%
For the distribution of the characteristic time of appearance, the
author explores exponential and truncated Gaussian distributions.
%
Moreover, for the distribution of the lifespans, the author chooses an
exponential distribution.
%
It is important to consider that the fraction of communicating
civilisations vary with the choice of the statistical distribution for
the time of appearance, as shown by \citet{balbi_impact_2018}.
%
For all analysed distributions, the author concludes that,
the detectable fraction corresponding to

for a mean
lifetime of 10$^9$ years and for a maximum radius for the detection of
the signal of 1000~ly, the detectable civilizations
are roughly the 10 per cent of the total, due to their spread in time and
their limited lifetimes.
%
In another related approach, \citet{vukotic_astrobiological_2012}
propose a probabilistic cellular automata (PCA) modelling. 
%
In this framework, a complex system is modelled by a lattice of cells
which evolve at discrete time steps, according to transition rules
that take into account for each cell the states of its neighbour
cells.
%
The authors implement a PCA model to a network of cells which
represent life complexity on a 2-dimensional region resembling the
Galactic Habitable Zone (GHZ),  an annular ring set between a minimal
radius of 6 kpc and a peak radius of 10 kpc. 
%
Within this framework, \citet{vukotic_astrobiological_2012} also make
several Monte Carlo simulations and analyse ensemble-averaged results.
%
The work aims at analysing the evolution of life, although it does not
account for the network of causal contacts among
technological civilisations.

@@

In this work we address the problem of the temporal and spatial structure of
the distribution of communicating civilisations.
%
We make no asumptions about, e.g., the origin of life, the development of
intelligence or the formation of habitable planets according to stellar type.
%
Instead, we assume monoparametric function families to model the appearance of
points in the disk of the Galaxy and over the time, which we call
\textit{nodes}.
%
These nodes represent the locations of ideal intelligent agents that are able
to receive and emit signals with perfect efficiency in all directions.
%
Then we analyse the network of causally connected nodes limited to have a
maximum separation representing the maximum distance a signal can travel with
an intensity above a fixed threshold.
%
A node is causally connected to other node if it is within their light cone.
%
We adopt as a definition that a light cone is the region in spacetime within
the surface generated by the light emanating from a given point in space in a
given period of time.
%
For example, two nodes separated by 100 light years which appear with a
difference in time of 80 years and last 50 years each, will be causally
connected for 30 years.
%
This connection is not, though, bidirectional.
%
In fact, the second node will see the first one for 30 years, but the first
node will never acknowledge the existence of the second.
%
This simple tought experiment exemplify the importance of the time variable to
analyse the structure of connections.
%
The model can be described with three parameters, namely, the mean separation
between the appearance of nodes, the mean lifetime of the nodes, and the
maximum separation to allow an effective causal contact.
%
The model presented here comprises the stochastic networks of constrained
causally connected nodes, together with the three parameters and additional
hypotheses.
%
Hereafter, we refeer to this model as the C3Net model, standing for
''Constrained Causally Conencted Network``.

We provide a framework to explore, through a suite of numerical simulations,
the parameter space of three unknown observables.
%
This allows to discuss possible scenarios and their consequences in terms of
the probability of making contacts.
%
The method we use for simulating a stochastic process is an approximation that
allows to study the behaviour of complex systems, by considering a sequence of
well defined discrete events.
%
The simulation is carried out by following all the variables that describe and
constitute the state of the system, and the evolution of the process is
described as a set of changes in that state.
%
In this context, an event produces a specific change in the state that can be
triggered by random variables that encode the stochastic nature of the physical
phenomenon.
%
For example, when a new contact is produced between two entities in the
simulated galaxy, the number of active communication lines is increased by one.
%
Also, if it is the first contact for that nodes, 
then the number of communicated nodes increases by two.
%
When a new node becomes active, the sistem has an increase of one
in the number of nodes, although the number of communcations does not
necessarily change.
%
The process then involves following the changes on the state of the system, defining
the initial and final states.
%
This is done by defining a method that allows to keep track of the time
progress in steps and maintaining a list of relevant events, i.e., the events
that produce a change in the vriables of interest.
%
With this method we abandon the frequentist approach of the Drake equation to
compute the number of civilisations, providing instead its statistical
distribution.
%
More importantly, we are interested on the probability of contacts, which
depend critically on the time variable.
%
This is an exploratory analysis that aims at developing a numerical tool to
discuss the different scenarios based on statistical heuristics.
%
The approach proposed here should be considered as a compromise between the
uncertainties of the frequentist estimations and the detailed recipes required
on the numerical simulations.
%
It is worth noticing that the model is not intended for a formal fit at this
stage due to the lack of data, but is can help to understand how unusual it
would be to actively search for intelligente signal for 50 years without
success.

This paper is organized as follows.
%
In Sec.~\ref{S_methods} we introduce the methods and discuss the candidate
distributions for the statistical aspects of the times involved in the
communication process.
%
We present our results in Sec~\ref{S_results}, with special emphasis on the
statistical distributions of the duration of causal contacts in one or both
directions and the distribution of time intervals for the waiting of the first
contact.
%
This quantities are considered as a function of the three simulation
parameters.
%
In Sec.~\ref{S_discussion} we discuss our results and future research
directions.   
%
In this work we use the International System of units recommended by the
International Astronomial Union, where the simbol for years is ''a''.

\begin{figure}[!t]
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   %
   \caption{
   %
Space--time diagrams showing a schematic representation of the different stages
in the development of a constrained causal contact node (\ceti{}).
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (E, left vertex), following an ``A'' type
event (``Awakening'') in which the node acquires the communication
capability.
%
The sphere of first contact (SFC) is the sphere centered on the emitter that
grows until its radius reaches the D$_{\rm max}$ distance, in which the
power of the signal would equal the detectability threshold.
%
In the Figure this sphere is represented by the left triangle of the region
$R$.
%
The surface of last contact (SLC) is another sphere that grows from a ``D''
event (``Doomsday'').
%
The region which is causally connected to the emitter is then limited by these
two spheres, and has the shape of a sphere or of a spherical shell,
depending on the time.
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b).
%
The receiver node E$_1$ can listen signals from the emitter node E$_2$, from
the ``Contact'' event (t=C$_{12}$) up to the ``Blackout'' event
(t=B$_{12}$).
%
Similarly, the receiver \ceti{} E$_2$ can listen signals from the emitter
\ceti{} E$_1$, from t=C$_{21}$ up to t=B$_{21}$. 
%
} \label{F_scheme}
%
\end{figure}

@@

Simulations are suitable tools to analyse systems that evolve with time and
involve randomness.
%
An advantage of a numerical approach is that it usually requires fewer assumptions
and simplifications, and can be applied to systems where analytical models are
hard or impossible to develop.
%
In particular, a suitable tool to model complex stochastic processes through
the changes in the state of a system is the discrete--event (hereafter, DE)
simulation approach.
%
A system described with the DE paradigm is characterised by a set of actors and
events.
%
Actors interact causally through a series of events on a timeline and process
them in chronological order \citep{ptolemaeus_system_2014,
chung_simulation_2003, ross_simulation_2012}.
%
Each event modify the variables that define the process, producing the
corresponding change in the state of the simulated system.
%
This method is well suited for the particular case of the diffusion of
intelligent signals in the galaxy and allows to explore several models easily.
%
We simulate the statistical properties of a set of points in space and time
that have a causal connection at light speed and are separated
by a maximum distance.
%
We refer to this nodes as ''Constrained Causal Contact Nodes``
(hereafter, \ceti{}).
%
We chose this generic name in order to stress the fact that in this analysis 
only the causal contact is considered, independently of any broadcasting or
lookout activity.
%
The system we propose is ideal, in the sense that it considers the special case
of a fully efficient node that emits and receives isotropically.
%
A causal contact node, considered as a broadcasting station that has the
ability to detect signals through an active search program, could have a
lesser than one efficiency factor, which is not included in the current
analysis.
%
Also, it is worth mentioning that this is a general approach, and not
necessarily a \ceti{} is the host of an intelligent civilization.
%
It can be associated with a planet where life has developed,
became intelligent, reached the skills required to find the right communication
channel, sustained a search and established a contact.
%
Alternative message processing entities could be considered, for example
interstellar beacons where life has ceased to exist but continue with its
emission, or communication stations established by probes or left by
intelligent beings \citep[see, e.g., ][]{peters_outer_2018,
barlow_galactic_2013}.
%
In principle, these strategies could affect our results since it would be
easier to configure a cluster of \cetis{} that spread in time.
%
However, we do not consider these speculative alternatives at this point.
%
For the purpose of the present analysis, only the communication capability is
relevant, since we study the causal contacts between the locations.
%
The system is defined by a number of actors that represent \cetis{} and appear
at different instants in time, generating events that produce meaningful
changes in the variables that describe the system, i.e., in the arrangement of
\cetis{} and their network of causal contacts.
%
For example, the appearance of a new \ceti{} in a region filled with a signal
emitted by another node, will increase the number of active \cetis{} and the
number of pairs of \cetis{} in causal contact.
%
Assuming some simple hypotheses, the discrete events method can be performed
taking into account a small number of variables, which allow to analyze the
variation of the results in the model parameter space.

In what follows, we outline the experimental setup adopted to estimate the
probabilities of causal contacts and several derived quantities.
%
This is done in terms of three independent parameters, namely, the mean time
span between the appearance of consecutive \cetis{}, $\left<\tau_a\right>$,
their mean lifetime $\left<\tau_s\right>$, and the maximum distance a signal
can be detected by another \ceti{} ($D_{\rm max}$).
%
Intuitively, the probability of the existence of causal contacts between pairs
of \cetis{} would be larger for smaller $\tau_a$ parameters, higher $\tau_s$ or
higher $D_{\rm max}$ parameter values.
%
We also propose theoretical distribution functions for both the lifespans
($\tau_s$) and the number of \cetis{} per unit time
\citep{maccone_evolution_2014, Sotos_biotechnology_2019}.
%
The later is related to the time span between the appearance of consecutive
\cetis{} (since when $\left<\tau_a\right>$ is shorter, it produces a greater
density of \cetis{}).
%
The analytical expresions for these distributions are set to a fixed law, as
discussed in Sec.~\ref{SS_PDF_shape}.

@@

The temporal structure that emerges from the experimental setup implies that, as a fundamental property of the simulations, a causal connection can be produced without 
requiring that the two nodes are active at the same time.
%
This property of the system arises as a consequence of the large spatial and temporal scales, where a message is transmitted at the (relatively small) light speed.
%
Although a \ceti{} could be active for a large enough period to transmit a message at large distances, the limited power of the message and the dilution that depends on the squared distance from the source imposes a detectability limit.
%
As a consequence of this limitation and of their finite lifetime, considered as the period between the acquisition and loss of communicating capacity, each \ceti{} will fill a spherical shell region of the galaxy, limited by two concentric spherical surfaces.
%
The leading front, or surface of first contact (SFC) grows from the
central \ceti{} until it reaches the maximum distance $D_{\rm max}$.
%
Following the end of the civilization, there is still a region which is filled with the emitted signals.
%
This approach has been also considered in other statistical models \citep[e.g., ][]{smith_broadcasting_2009, grimaldi_signal_2017, Grimaldi2018}.
%
The trailing front, or surface of last contact (SLC) also grows from the central \ceti{}, with a delay with respect to the SFC equivalent to the lifetime of the node, and produces a spherical shell region.
%
Any other \ceti{} within this region will be aware of the originating node, even if it has disappeared before the time of contact.
%
This region will grow if the surface of first contact has not yet
reached the maximum distance $D_{\rm max}$, and will shrink after a
D-type event until the surface of last contact reaches $D_{\rm max}$, producing as a result the loss of all signal from the central \ceti{}.
%
In our approach, we consider a model galaxy where the width of the disk is negligible with respect to the radius of the disk.
%
In the 2D simulation only the intersection of the communicating spherical shells with the plane of the galaxy is relevant, and produce the corresponding circles or rings for the filled spheres or annular regions of the spherical shells, respectively.
%
The initially growing communicating sphere is shown over space--time diagrams, where time is represented on the horizontal direction, and space is represented in the vertical direction.
%
In the Fig.~\ref{F_scheme} the two emitters in panel (b), E$_1$ and E$_2$, reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the distance and time lag between the awakening events, the maximum distance that a signal can reach and the time period in which each emitter is active.
